Constructing Quotient Inductive-Inductive Types

### 2018-10-30 paper.tex
stuff

A quotient inductive-inductive type (QIIT) can be seen as a
multi-sorted algebraic theory where sorts can be indexed over each
other. An example of a QIIT is the following well-typed (intrinsic)
syntax of a small type theory.
\begin{alignat*}{5}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\Set \\
  & \cdot && : \Con \\
  & \rhd && : (\mathit{\Gamma}:\Con)\ra\Ty\,\mathit{\Gamma}\ra\Con \\
  & \U && : (\mathit{\Gamma}:\Con)\ra\Ty\,\mathit{\Gamma} \\
  & \Sigma && : (\mathit{\Gamma}:\Con)\ra(A:\Ty\,\mathit{\Gamma})\ra\Ty\,(\mathit{\Gamma}\rhd A)\ra\Ty\,\mathit{\Gamma} \\
  & \mathsf{eq} && : (\mathit{\Gamma}:\Con)\ra(A:\Ty\,\mathit{\Gamma})\ra(B:\Ty\,(\mathit{\Gamma}\rhd A))\ra (\Gamma\rhd A \rhd B = \Gamma\rhd \Sigma\,\Gamma\,A\,B)
\end{alignat*}
It has two sorts: contexts ($\Con$) and types ($\Ty$). The latter is
indexed over the former: to talk about a type we need to say which
context it lives in. There is an empty context $\cdot$ and context
extension $\rhd$ which takes a context and a type in that context and
returns the extended context. Note that we cannot turn this QIIT into
two inductive types defined one after the other because the
$\Con$-constructor $\rhd$ refers to $\Ty$, hence $\Con$ and $\Ty$ must
be defined \emph{at the same time}. There is a constructor for a base
type $\U$ in any context and a constructor for $\Sigma$ types. This
takes a context (this could be made an implicit parameter), a type $A$
in that context and a type in the context extended by $A$ and returns
a type in the original context. The third argument of $\Sigma$ shows a
pattern which does not appear in inductive types or indexed inductive
types: a constructor refers to a previous constructor (in our case
$\rhd$). Finally, there is an equality constructor which states the
unusual equality (included here for illustration) that extending a
context twice is the same as extending by a $\Sigma$ type. Equality
constructors can only target one of the sorts. The constructor
$\mathsf{eq}$ quotients contexts so that for any $\Gamma$, $A$ and $B$
it becomes impossible to distinguish $(\Gamma\rhd A \rhd B)$ and
$(\Gamma\rhd \Sigma\,\Gamma\,A\,B)$ -- this is ensured by the
eliminator of the QIIT as shown below. The equality constructor also
refers to previous constructors. This small example can be extended to
the full syntax of type theory as shown in \cite{ttintt}.

@@

We also have a dependent {\em function space with small domain}. This
can be used to add inductive arguments to operators in a signature,
for example the argument of sucessor for natural numbers. We have
constructors for $\Pi$, a categorical application rule and
substitution laws.
\begin{alignat*}{5}
  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\rhd\El\,a)\ra\Ty\,\Gamma \\
  & \APP && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra\Tm\,(\Gamma\rhd \El\,a)\,B \\
  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma^\uparrow]) \\
  & {\APP[]} && : (\APP\,t)[\sigma^\uparrow] = \APP\,(t[\sigma])
\end{alignat*}
By restricting the domain to be small, we can only write strictly
positive operators (we cannot write $\Pi\,(\Pi\,a\,B)\,C$ because the
first argument of $\Pi$ needs to be small but $(\Pi\,a\,B)$ itself is
large). There is no lambda, therefore this function type only has
neutral elements. We define the non-dependent function space by the
following abbreviation.
\[
(a:\Tm\,\Gamma\,\U)\Ra(B:\Ty\,\Gamma) :\Ty\,\Gamma :\equiv \Pi\,a\,(B[\wk])
\]
$\blank\app \blank$ is left-associative, $\blank\Ra\blank$ is
right-associative. The usual application can be defined as below.
\[
 (t:\Tm\,\Gamma\,(\Pi\,a\,B))\app\,(u : \Tm\,\Gamma\,(\El\,a)):\Tm\,\Gamma\,(B[\lb u\rb]) :\equiv (\APP\,t)[\lb u \rb]
\]

>>>>

A quotient inductive-inductive type (QIIT) can be seen as a
multi-sorted algebraic theory where sorts can be indexed over each
other. An example of a QIIT is the following well-typed (intrinsic)
syntax of a small type theory.
\begin{alignat*}{5}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\Set \\
  & \cdot && : \Con \\
  & \rhd && : (\mathit{\Gamma}:\Con)\ra\Ty\,\mathit{\Gamma}\ra\Con \\
  & \iota && : (\mathit{\Gamma}:\Con)\ra\Ty\,\mathit{\Gamma} \\
  & \Sigma && : (\mathit{\Gamma}:\Con)\ra(A:\Ty\,\mathit{\Gamma})\ra\Ty\,(\mathit{\Gamma}\rhd A)\ra\Ty\,\mathit{\Gamma} \\
  & \mathsf{eq} && : (\mathit{\Gamma}:\Con)\ra(A:\Ty\,\mathit{\Gamma})\ra(B:\Ty\,(\mathit{\Gamma}\rhd A))\ra (\Gamma\rhd A \rhd B = \Gamma\rhd \Sigma\,\Gamma\,A\,B)
\end{alignat*}
It has two sorts: contexts ($\Con$) and types ($\Ty$). The latter is
indexed over the former: to talk about a type we need to say which
context it lives in. There is an empty context $\cdot$ and context
extension $\rhd$ which takes a context and a type in that context and
returns the extended context. Note that we cannot turn this QIIT into
two inductive types defined one after the other because the
$\Con$-constructor $\rhd$ refers to $\Ty$, hence $\Con$ and $\Ty$ must
be defined \emph{at the same time}. There is a constructor for a base
type $\iota$ in any context and a constructor for $\Sigma$ types. This
takes a context (this could be made an implicit parameter), a type $A$
in that context and a type in the context extended by $A$ and returns
a type in the original context. The third argument of $\Sigma$ shows a
pattern which does not appear in inductive types or indexed inductive
types: a constructor refers to a previous constructor (in our case
$\rhd$). Finally, there is an equality constructor which states the
unusual equality (included here for illustration) that extending a
context twice is the same as extending by a $\Sigma$ type. Equality
constructors can only target one of the sorts. The constructor
$\mathsf{eq}$ quotients contexts so that for any $\Gamma$, $A$ and $B$
it becomes impossible to distinguish $(\Gamma\rhd A \rhd B)$ and
$(\Gamma\rhd \Sigma\,\Gamma\,A\,B)$ -- this is ensured by the
eliminator of the QIIT as shown below. The equality constructor also
refers to previous constructors. This small example can be extended to
the full syntax of type theory as shown in \cite{ttintt}.

@@

We also have a dependent {\em function space with small domain}. This
can be used to add inductive arguments to operators in a signature,
for example the argument of sucessor for natural numbers. We have
constructors for $\Pi$, a categorical application rule and
substitution laws.
\begin{alignat*}{5}
  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\rhd\El\,a)\ra\Ty\,\Gamma \\
  & \APP && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra\Tm\,(\Gamma\rhd \El\,a)\,B \\
  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma^\uparrow]) \\
  & {\APP[]} && : (\APP\,t)[\sigma^\uparrow] = \APP\,(t[\sigma])
\end{alignat*}
By restricting the domain to be small, we can only write strictly
positive operators (we cannot write $\Pi\,(\Pi\,a\,B)\,C$ because the
first argument of $\Pi$ needs to be small but $(\Pi\,a\,B)$ itself is
large). We omit lambda abstraction, because there is no essential use
for it in signatures, and it follows that this function type only has
neutral elements. We define the non-dependent function space by the
following abbreviation.
\[
(a:\Tm\,\Gamma\,\U)\Ra(B:\Ty\,\Gamma) :\Ty\,\Gamma :\equiv \Pi\,a\,(B[\wk])
\]
$\blank\app \blank$ is left-associative, $\blank\Ra\blank$ is
right-associative. The usual application can be defined as below.
\[
 (t:\Tm\,\Gamma\,(\Pi\,a\,B))\app\,(u : \Tm\,\Gamma\,(\El\,a)):\Tm\,\Gamma\,(B[\lb u\rb]) :\equiv (\APP\,t)[\lb u \rb]
\]
